---
title: "Having fun with the Canadian road network"
output: html_document
date: "2022-11-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
#library(OpenStreetMap)
library(sf)
```

Canada has much information online, so we should be able to hack something together... Some sources of information:

- [Road network file](https://www150.statcan.gc.ca/n1/en/catalogue/92-500-X)
- [National road network](https://open.canada.ca/data/en/dataset/3d282116-e556-400c-9306-ca1a3cada77f)
- A more general [site](https://freegisdata.rtwilson.com/) with a myriad of links to various GIS resources.

Let us try [this one](https://www12.statcan.gc.ca/census-recensement/2011/geo/RNF-FRR/index-s-eng.cfm?year=22), see if it works. Note that they mention a [reference guide](https://www150.statcan.gc.ca/n1/pub/92-500-g/92-500-g2022001-eng.htm). It is always a good idea to download that, because you can expect that things may get a bit complicated...

In terms of format, let's go with Shapefile (.shp). We get a ~350MB zip file that unzips to a 1.7GB folder. Unfortunately, that's common: geographic files are often large... 

```{r}
America_roads <- st_read("groads-v1-americas-shp/gROADS-v1-americas.shp")
```

```{r}
CAN_roads <- st_read("lrnf000r22a_e/lrnf000r22a_e.shp")
```


```{r}
sort(unique(CAN_roads$TYPE))
CAN_hwy = CAN_roads %>%
  filter(TYPE == "HWY")
rm(CAN_roads)
plot(CAN_hwy)
```



