---
title: "The Manitoba part of the global air transportation network"
author: "Julien Arino"
date: "2024-11-05"
output: html_document
---

```{r setup, include=TRUE, message=FALSE, echo=FALSE}
# By default, we won't show code chunks, just use echo=TRUE or specific lines
# to show them when needed
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      fig.width = 10,
                      fig.height = 8)
# Load/install required libraries. Note the use of Ncpus to speed up the process,
# when compiling packages is required (typically linux)
required_libs = c("dplyr", 
                  "GA",
                  "geodata", 
                  "ggraph", 
                  "igraph",
                  "kableExtra",
                  "knitr",
                  "readr", 
                  "sqldf")
for (l in required_libs) {
    if (!require(l, character.only = TRUE)) {
        install.packages(l, Ncpus = (parallel::detectCores()-1))
        library(l, character.only = TRUE)
    }
}
```

# Word of warning (for upcoming assignments)

I don't want to see exact replicates of this document. Here, you will note that I am "riffing", exploring potential ways to represent the information and discussing this. In your assignments, this is not something you would typically do. 
Explain what you are doing, but get to the point, don't "riff" around without a purpose.

# Load the data
We load two data files: `air_travel_MB.csv` contains the network description and `airport_codes_CAN.csv`, which has some information about all airfields in Canada.

The data in air_travel_MB is pre-processed. It represents an extreme simplification of a much more complete dataset giving, for a given time period, the total number of travellers between 3734 airports worldwide along trips including up to 5 intermediate stops. 

What I did is to keep only the airports in Manitoba. Any airport outside of Manitoba is set as `RoW` (rest of the world). Also, the existing multi-leg trips within Manitoba are "decomposed": if the data had 10 people flying, say, YWG$\to$YQD$\to$YTH (Winnipeg to The Pas to Thompson), these would appear as 10 people flying from YWG to YQD and 10 people flying from YQD to YTH. 

Flights to and from RoW are decomposed the same way as flights within Manitoba, increasing the number of trips between RoW and YWG. Indeed, 5 passengers flying RoW$\to$YWG$\to$YQD and 5 passengers flying RoW$\to$YWG$\to$YBR (Brandon) will show as 10 passengers flying RoW to YWG, together with 5 passengers flying YWG$\to$YQD and 5 passengers flying YWG$\to$YBR.

First, here is what the flight data looks like. (I am using `knitr::kable` to make the table more readable.)

```{r load-air-travel-file}
air_travel_MB <- read_delim("air_travel_MB.csv", 
                            delim = ",", escape_double = FALSE,
                            trim_ws = TRUE)
knitr::kable(head(air_travel_MB))
```

The airport data, filtered to only retain MB locations, looks as follows.

```{r load-airports-file}
airport_codes_MB <- read_csv("airport_codes_CAN.csv") %>%
  filter(ISO_Region == "CA-MB")
knitr::kable(airport_codes_MB) %>%
  kable_styling("striped", full_width = F) %>% 
 scroll_box(width = "800px", height = "400px")
```


# Data wrangling remark
Note that there is an alternative to `dplyr` (the package that allows us to use the pipe `%>%`); it is indeed possible to treat data frames as SQL tables and run SQL queries on them. This may make more sense to some of you. For illustration, to create the table of movements, I used the following command (which can run, since it is not destructive of what we have already done).


```{r example-sqldf, echo=TRUE}
query = "SELECT orig,dest,SUM(vol) AS vol
FROM air_travel_MB
GROUP BY orig,dest
ORDER BY orig,dest"
air_travel_MB = sqldf::sqldf(query)
```

Finish preparing the data: the following code (chunk not shown) adds the latitude and longitudes of all the airports in the `air_travel_MB` data set. Note that we give `RoW` a location: we will show it as an additional vertex in the graph. (I picked somewhere in Ontario, close to but not in Manitoba.)

```{r add-lat-lon-info-to-airports}
IATA_codes_MB_in_data = 
  data.frame(IATA = sort(unique(union(air_travel_MB$orig, air_travel_MB$dest))))
lon = c()
lat = c()
for (i in 1:dim(IATA_codes_MB_in_data)[1]) {
  idx = which(airport_codes_MB$IATA_Code == IATA_codes_MB_in_data$IATA[i])
  if (length(idx) == 0) {
    lon = c(lon, -90)
    lat = c(lat, 49.5)
  } else {
    lon = c(lon, airport_codes_MB$Longitude[idx])
    lat = c(lat, airport_codes_MB$Latitude[idx])
  }
}
IATA_codes_MB_in_data$lon = lon
IATA_codes_MB_in_data$lat = lat
kable(head(IATA_codes_MB_in_data))
```




We are now in a position to make the graph and plot it in simple form. We add the volume as the weight of the arcs.


```{r plot-graph-1}
G = graph_from_data_frame(air_travel_MB)
V(G)$x = lon
V(G)$y = lat

plot(G, edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```

We also make a subgraph without `RoW` to focus on the properties really specific to Manitoba.


```{r plot-MB-graph-without-RoW-1}
G_MB = induced_subgraph(G, V(G)$name[V(G)$name != "RoW"])
plot(G_MB, edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```

# Making nicer plots
Before we explore the properties of the graph, let us spend a little time on visualisation. This is not to be neglected: nice plots help more than you think.

The next piece of code does the following:

- Redefine the output size (does it once for all, need to modify if need be)
- Uses the library `geodata` to read in the shape file for Canada
- Selects Manitoba from the Canada shape file
- Plots Manitoba
- Plots the graph


```{r plot-graph-with-MB-shape}
Canada <- geodata::gadm(country='CAN', level=1, path = ".")
MB <- Canada[Canada$NAME_1 == "Manitoba",]
plot(MB)
plot(G, add = TRUE, rescale = FALSE,
     edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```

That box around the plot is a little distracting, let's remove it.

```{r plot-graph-with-MB-shape-2}
plot(MB, box = FALSE, axes = FALSE)
plot(G, add = TRUE, rescale = FALSE,
     edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```


Let us add the degree information as the size of the vertices. First, on the graph with rest of the world connection information..


```{r}
plot(MB, box = FALSE, axes = FALSE)
V(G)$size = degree(G, mode = "out")*10
plot(G, add = TRUE, rescale = FALSE,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```

Then on the "purely MB" graph.

```{r}
plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = degree(G_MB)*5
plot(G_MB, add = TRUE, rescale = FALSE,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```

So clearly, Winnipeg (YWG) and Thompson (YTH) have the highest degree.
Let us confirm this by looking at the degree distribution.

```{r}
sort(degree(G_MB), decreasing = TRUE)
```
So, within the MB graph, Winnipeg dwarfs all other vertices in terms of connections to other locations. Note this is the degree in the underlying undirected graph, if we choose an orientation, we get a slightly different picture.

```{r}
sort(degree(G_MB, mode = "in"), decreasing = TRUE)
sort(degree(G_MB, mode = "out"), decreasing = TRUE)
dG_MB_in = degree(G_MB, mode = "in")
dG_MB_out = degree(G_MB, mode = "out")
diff_d = names(which(dG_MB_out != dG_MB_in))
```

It looks like the degree is usually the same regardless of whether we consider in- or out-degree. 
The only `r length(diff_d)` airports that have different in- and out-degree are `r `diff_d`, that is, `r airport_codes_MB$Name[airport_codes_MB$IATA_Code %in% diff_d]`.


Now let us take a look at the arcs. Maybe change their thickness to reflect the volumes along the arcs?


```{r}
plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = degree(G_MB)*5
plot(G_MB, add = TRUE, rescale = FALSE, edge.width = E(G_MB)$vol,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```


My guess: some of the arcs are so massive that this is all we can see. So let us renormalise the values of the weights. First, what do they look like now?


```{r}
sort(E(G)$vol)
```


So, yes, clearly a bit too much variation. If we scale linearly by multiplying by a constant, we will most likely lose the information on the small weights. Just to check that we are intuiting correctly, though, let us try once doing this.


```{r}
plot(MB, box = FALSE, axes = FALSE)
V(G)$size = degree(G)*5
plot(G, add = TRUE, rescale = FALSE, edge.width = E(G)$vol*0.001,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```

```{r}
plot(MB, box = FALSE, axes = FALSE)
V(G)$size = degree(G)*5
plot(G, add = TRUE, rescale = FALSE, edge.width = E(G)$vol*0.0005,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```


What we suspected, indeed: to make the arcs with the largest volume not too large, we had to essentially get rid of the weight of the arcs with small volumes. So what we want to do is to process the values through a function that would make the smaller values larger relative to the larger ones or the larger values closer to the smaller one, then scale the result. For this, it is good to look at what the values look like.


```{r}
plot(sort(E(G)$vol), type = "b")
```

One way around this could be to use a so-called sigmoid function. This is a function of the following form:
$$
f(y) = \dfrac{1}{1+e^{-cy}},
$$
where $\mathbb{R}\ni c>0$ is a parameter.
This function has the following properties: $\lim_{y\to-\infty}f(y)=0$, $\lim_{y\to\infty}f(y)=1$ and $f(0)=1/2$. By playing with the value of $c$, one can set the beahviour of $f$.
Let us transform the values of the arc weights using such a function, testing how different values of $c$ influence the value of the points.
Because we are only transforming positive values, let us work with the positive half of the sigmoid.

In the plot below, the red points are the original values, unscaled, while the different colours are the transformed values.
(The `viridis` palette used starts with yellow and goes towards green and then purple.)


```{r}
sigmoid = function(x, c) {
  # Normalise input values to [0,1] for convenience
  x = x-min(x)
  x = x / max(x)
  # We only use half of the function, so normalise results
  # to the interval [0,1]. 
  if (min(x)<0) {
    return(NULL)
  }
  values = 1/(1+exp(-c*x))
  values = values - min(values)
  values = values / max(values)
  return(values)
}

values_c = c(0.1, 1, 10, 100, 1000)
values_x = 1:length(E(G))
values_y = sort(E(G)$vol)
values_y = (values_y-min(values_y))
values_y = values_y / max(values_y)
my_palette = viridis::viridis(length(values_c))

# First, we plot the "data" (scaled to the [0,1] range)
plot(values_x, values_y, 
     type = "b", col = "red",
     ylim = c(0, 1))
for (c in values_c) {
    idx = which(c == values_c) 
    tmp = sigmoid(values_y, c)
    lines(values_x, tmp, 
          type = "b", col = my_palette[idx])
}
```

Somewhere between the last two looks quite nice. Let's dig a bit more around, see if we can be a bit better. We can use a genetic algorithm to find the value of $c$ that makes the transformed values as close as possible to a line from the minimum $(x_m,y_m)$ to the maximum value $(x_M,y_M)$.
The line has equation
$$
y = \frac{y_M - y_m}{x_M - x_m}(x - x_m) + y_m
$$
To do this, we define an error function that we want to minimise and run the genetic algorithm, as shown below.
```{r find-best-sigmoid, echo=TRUE}
error_transformation = function(data_x, data_y, c) {
  x_m = min(data_x)
  x_M = max(data_x)
  y = 1 / (x_M - x_m) * (data_x - x_m)
  transformed_y = sigmoid(data_y, c)
  error = sum((transformed_y-y)^2)
  return(error)
}
GA = ga(type = "real-valued", 
        fitness = 
          function(x) -error_transformation(values_x, 
                                            values_y, 
                                            x),
        lower = c(100),
        upper = c(1000),
        suggestions = c(150))
values_c = as.numeric(GA@solution)
```
We find a value of $c\simeq$ `r round(values_c,3)`. Using this value, the transformed values take the form below.
```{r}
# First, we plot the "data" (scaled to the [0,1] range)
plot(values_x, values_y,
     type = "b", col = "red",
     ylim = c(0, 1))
for (c in values_c) {
    idx = which(c == values_c)
    tmp = sigmoid(values_y, c)
    lines(values_x, tmp, 
          type = "b", 
          col = my_palette[idx])
}
x_m = min(values_x)
x_M = max(values_x)
y = 1 / (x_M - x_m) * (values_x - x_m)
lines(values_x, y/max(y), type = "b")
```

We most likely will have to multiply by a factor, so we compute the values in advance.


```{r}
edge_widths = sigmoid(as.numeric(E(G)$vol), values_c)

plot(MB, box = FALSE, axes = FALSE)
V(G)$size = degree(G)*5
plot(G, add = TRUE, rescale = FALSE, 
     edge.width = 10*edge_widths, type = "b",
     edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```

Well, after all this, I am not convinced we achieved better than we had done above with just a linear scaling. So when we want a a good plot, we'll come back to that solution.


```{r}
plot(MB, box = FALSE, axes = FALSE)
V(G)$size = degree(G)*5
plot(G, add = TRUE, rescale = FALSE, edge.width = E(G)$vol*0.0005,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```


```{r}
plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = degree(G_MB)*5
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```


## Eccentricities and related measures


```{r}
distances(G_MB)
```



```{r}
distances(G_MB, mode = "out")
```





```{r}
eccentricity(G_MB)
eccentricity(G_MB, mode = "in")
eccentricity(G_MB, mode = "out")
writeLines("Radius (all/in/out):")
radius(G_MB)
radius(G_MB, mode = "in")
radius(G_MB, mode = "out")
writeLines("Diameter (all/in/out:")
diameter(G_MB)
max(distances(G_MB, mode = "in"))
max(distances(G_MB, mode = "out"))
```


## Central points, centre, periphery
For simplicity, I only show the out-degree case. And just to show how easily it's done: let's colour the vertices in the centre red and those in the periphery blue.


```{r}
e_out = eccentricity(G_MB, mode = "out")
writeLines("Central points:")
V(G_MB)[which(e_out == min(e_out))]$name
V(G_MB)[which(e_out == min(e_out))]$color = "red"
writeLines("Periphery:")
V(G_MB)[which(e_out == max(e_out))]$name
V(G_MB)[which(e_out == max(e_out))]$color = "blue"

plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = degree(G_MB)*5
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```


## Girth
Recall that `girth` in `igraph` assumes the graph is undirected.


```{r}
girth(G)
girth(G_MB)
```



## Density
What fraction of possible arcs/edges are present?


```{r}
edge_density(G)
edge_density(G_MB)
```



# Connectedness and related
## Connectedness


```{r}
is.connected(G_MB, "weak")
is.connected(G_MB, "strong")
```




So there is a single strongly connected component.
## Articulation points
The function assumes the graph is undirected, even in the directed case.


```{r}
articulation_points(G_MB)
```

So removing Winnipeg or Thompson disconnects the graph.

```{r}
G_disconnected = induced_subgraph(G_MB, V(G_MB)$name[V(G_MB)$name != "YTH"])
is_connected(G_disconnected)

comps <- components(G_disconnected)$membership
colbar <- rainbow(max(comps)+1)
V(G_disconnected)$color <- colbar[comps+1]

plot(MB, box = FALSE, axes = FALSE)
V(G_disconnected)$size = rep(50,length(V(G_disconnected)))
plot(G_disconnected, add = TRUE, rescale = FALSE, 
     edge.width = E(G_disconnected)$vol*0.001,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```


# Cliques


```{r}
lc = largest_cliques(G_MB)
lc
vert_largest_clique = lc[[1]]

V(G_MB)$color = "blue"
V(G_MB)[vert_largest_clique]$color = "red"

plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(50,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_disconnected)$vol*0.001,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
```

# Centrality
## Degree-centrality


```{r}
degree(G_MB, mode = "out")
degree(G_MB, mode = "in")
degree(G_MB)
distrib_outdegree = degree_distribution(G, mode = "out")
distrib_indegree = degree_distribution(G, mode = "in")
distrib_degree = degree_distribution(G)
```





```{r}
options(repr.plot.width=10, repr.plot.height=10)

distrib_outdegree = degree_distribution(G_MB, mode = "out")
distrib_indegree = degree_distribution(G_MB, mode = "in")
distrib_degree = degree_distribution(G_MB)

y_max = max(max(distrib_indegree, distrib_outdegree), distrib_degree)
plot(distrib_degree, type = "b", lwd = 2, col = "black", ylim = c(0, y_max))
points(distrib_outdegree, type = "b", lwd = 2, col = "blue")
points(distrib_indegree, type = "b", lwd = 2, col = "red")
```

# K-nearest neighbours
For info, the return values of the function `knn` are
- `knn`: A numeric vector giving the average nearest neighbor degree for all vertices in vids.
- `knnk`: A numeric vector, its length is the maximum (total) vertex degree in the graph. The first element is the average nearest neighbor degree of vertices with degree one, etc.


```{r}
knn(G_MB, mode = "all")
knn(G_MB, mode = "in")
knn(G_MB, mode = "out")
```



Let us plot the values in the graph, say, for the in-knn.


```{r}
options(repr.plot.width=15, repr.plot.height=15)

knn_in = knn(G_MB, mode = "in")$knn
knn_in_bar = sort(unique(round(knn_in)))
colbar <- viridis::viridis(max(ceiling(knn_in)))
V(G_MB)$color <- colbar[round(knn_in)]

plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)
DescTools::ColorLegend(x = "right", col = colbar, 
                       width = 0.5, inset = 0.1,
                       labels = sprintf("%d", knn_in_bar))
```


# Coreness


```{r}
coreness(G_MB, "all")
coreness(G_MB, "in")
coreness(G_MB, "out")
```

Let us look at the out-coreness, for instance, and colour the vertices as a function of that.


```{r}
cores = coreness(G_MB, "out")
colbar <- viridis::viridis(max(cores))
V(G_MB)$color <- colbar[cores]

plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
    edge.arrow.size = 0.5, edge.arrow.width = 0.75)
legend("bottomright", 
       legend = sprintf("%d", sort(unique(cores))),
       pch = rep(19, length(unique(cores))),
       pt.cex = rep(5, length(unique(cores))),
       cex = 2,
       box.lty=0,
       col = colbar[sort(unique(cores))])
```


# Betweenness


```{r}
betweenness(G)
betweenness(G_MB)
```




```{r}
betweenness_MB = betweenness(G_MB)
# shift numbers by 1 because the second command fails when there are zeros..
colbar <- viridis::viridis(max(ceiling(betweenness_MB))+1)
V(G_MB)$color <- colbar[round(betweenness_MB)+1]
# For the label of the colour bar
colbar_str = sprintf("%d", range(unique(round(betweenness_MB))))

plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)
DescTools::ColorLegend(x = "right", col = colbar, 
                       width = 0.5, inset = 0.1,
                       labels = colbar_str)
```


# Closeness
First, the unweighted case, i.e., using the geodesic distance.


```{r}
round(closeness(G_MB, mode = "all"),5)
round(closeness(G_MB, mode = "in"),5)
round(closeness(G_MB, mode = "out"),5)
```



We can also compute closeness using weighted arcs, i.e., using the travel volumes. Geodesic distance shown for comparison. We save the results for future manipulation.


```{r}
# Unweighted case
cl_MB_U = closeness(G_MB, mode = "out")
cl_MB_U
# Weighted case
cl_MB_W = closeness(G_MB, mode = "out", weights = E(G_MB)$vol)
cl_MB_W
```








To prepare for plots, let us look at the ranges of values.


```{r}
options(repr.plot.width=10, repr.plot.height=10)

range(cl_MB_U)
plot(sort(cl_MB_U))
range(cl_MB_W)
plot(sort(cl_MB_W))
```


So if we multiply the first values by 10,000 and the second ones by 1e7, we should get good ranges of variation.


```{r}
range(cl_MB_U * 1e4)
range(cl_MB_W * 1e7)
```



Indeed, so that's what we do.


```{r}
options(repr.plot.width=15, repr.plot.height=15)

cl_MB_U_tmp = cl_MB_U * 1e4
Delta_cl = ceiling(max(cl_MB_U_tmp)-min(cl_MB_U_tmp))
colbar <- viridis::viridis(Delta_cl+1)
V(G_MB)$color <- colbar[round(cl_MB_U_tmp-min(cl_MB_U_tmp))+1]
# For the label of the colour bar
colbar_str = sprintf("%f", range(cl_MB_U_tmp))

plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)
DescTools::ColorLegend(x = "right", col = colbar, 
                       width = 0.5, inset = 0.1,
                       labels = colbar_str)
```


```{r}
cl_MB_W_tmp = cl_MB_W * 1e7
Delta_cl = ceiling(max(cl_MB_W_tmp)-min(cl_MB_W_tmp))
colbar <- viridis::viridis(Delta_cl+1)
V(G_MB)$color <- colbar[round(cl_MB_W_tmp-min(cl_MB_W_tmp))+1]
# For the label of the colour bar
colbar_str = sprintf("%f", range(cl_MB_W_tmp))

plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)
DescTools::ColorLegend(x = "right", col = colbar, 
                       width = 0.5, inset = 0.1,
                       labels = colbar_str)
```

One interesting remark here: closeness radically changes when considering the unweighted and weighted version. Just to confirm this, let us plot the two graphs side by side.


```{r}
par(mfrow=c(1,2))

# Unweighted MB only
Delta_cl = ceiling(max(cl_MB_U_tmp)-min(cl_MB_U_tmp))
colbar <- viridis::viridis(Delta_cl+1)
V(G_MB)$color <- colbar[round(cl_MB_U_tmp-min(cl_MB_U_tmp))+1]
# For the label of the colour bar
colbar_str = sprintf("%f", range(cl_MB_U_tmp))
plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)

# Weighted MB only
Delta_cl = ceiling(max(cl_MB_W_tmp)-min(cl_MB_W_tmp))
colbar <- viridis::viridis(Delta_cl+1)
V(G_MB)$color <- colbar[round(cl_MB_W_tmp-min(cl_MB_W_tmp))+1]
# For the label of the colour bar
colbar_str = sprintf("%f", range(cl_MB_W_tmp))
plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)
```

So Thompson (YTH) is has higher closeness centrality than Winnipeg (YWG) when passenger volumes are considered. Does the situation change if we recall that Winnipeg is quite connected to the rest of the world? We prepare the computations and the plots as we had done earlier.


```{r}
# Unweighted case
cl_MB_RoW_U = closeness(G, mode = "out")
range(cl_MB_RoW_U)
# Weighted case
cl_MB_RoW_W = closeness(G, mode = "out", weights = E(G)$vol)
range(cl_MB_RoW_W)
```



Ranges are quite similar to those earlier, so we can probably just do the same as before. We will plot the 4 figures this time. First row is as before, second row is with rest of the world information thrown in.


```{r}
par(mfrow=c(2,2))

cl_MB_RoW_U_tmp = cl_MB_RoW_U * 1e4
cl_MB_RoW_W_tmp = cl_MB_RoW_W * 1e7

# Unweighted MB only
Delta_cl = ceiling(max(cl_MB_U_tmp)-min(cl_MB_U_tmp))
colbar <- viridis::viridis(Delta_cl+1)
V(G_MB)$color <- colbar[round(cl_MB_U_tmp-min(cl_MB_U_tmp))+1]
# For the label of the colour bar
colbar_str = sprintf("%f", range(cl_MB_U_tmp))
plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)

# Weighted MB only
Delta_cl = ceiling(max(cl_MB_W_tmp)-min(cl_MB_W_tmp))
colbar <- viridis::viridis(Delta_cl+1)
V(G_MB)$color <- colbar[round(cl_MB_W_tmp-min(cl_MB_W_tmp))+1]
# For the label of the colour bar
colbar_str = sprintf("%f", range(cl_MB_W_tmp))
plot(MB, box = FALSE, axes = FALSE)
V(G_MB)$size = rep(75,length(V(G_MB)))
plot(G_MB, add = TRUE, rescale = FALSE, 
     edge.width = E(G_MB)$vol*0.001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)

# Unweighted MB with RoW
Delta_cl = ceiling(max(cl_MB_RoW_U_tmp)-min(cl_MB_RoW_U_tmp))
colbar <- viridis::viridis(Delta_cl+1)
V(G)$color <- colbar[round(cl_MB_RoW_U_tmp-min(cl_MB_RoW_U_tmp))+1]
# For the label of the colour bar
colbar_str = sprintf("%f", range(cl_MB_RoW_U_tmp))
plot(MB, box = FALSE, axes = FALSE)
V(G)$size = rep(75,length(V(G)))
plot(G, add = TRUE, rescale = FALSE, 
     edge.width = E(G)$vol*0.0001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)

# Weighted MB with RoW
Delta_cl = ceiling(max(cl_MB_RoW_W_tmp)-min(cl_MB_RoW_W_tmp))
colbar <- viridis::viridis(Delta_cl+1)
V(G)$color <- colbar[round(cl_MB_RoW_W_tmp-min(cl_MB_RoW_W_tmp))+1]
# For the label of the colour bar
colbar_str = sprintf("%f", range(cl_MB_RoW_W_tmp))
plot(MB, box = FALSE, axes = FALSE)
V(G)$size = rep(75,length(V(G)))
plot(G, add = TRUE, rescale = FALSE, 
     edge.width = E(G)$vol*0.0001,
     edge.arrow.size = 0.5, 
     edge.arrow.width = 0.75)
```



Actually, things do not change, except that RoW becomes important as well.